---
fontsize: 12pt
linestretch: 2
output:
  pdf_document: default
  html_document: default
---

**Aim 1:**

We will consider outcomes as the change in each of the variables in table 3, as well as cortical thickness, over 1 year. Using multiple linear regression, we will model these continuous outcomes predicted by all inflammatory markers - (IL-6, TNF- alpha,MCP-1,Eotaxin-1) (all continuous) at baseline and as their difference over a year. Unless model assessment shows poor fit, we will assume linear relationships between outcomes and predictors. We will explore variable distributions to see whether data transformation (e.g. log-transforming right-skewed variables) is necessary. Missing values and outliers will be flagged and potentially removed, for example if an observation has multiple markers with values further than 2 SDs from the mean. We will assess covariates (age, sex, CV risk, immunological health history conditions, APOE genotype, baseline outcome values) and determine whether they are confounders (causally affecting the predictors and outcomes), precision variables (causally affecting the outcome), or mediators (variables through which the predictor impacts the outcome), using both domain knowledge and regression result assessment. We will check whether covariates are highly correlated with each other ($\rho$ > .7). If two precision variables are strongly correlated, only the more predictive one will be kept in the model. We will test the overall effect of the markers on the outcomes in the presence of age, sex, and precision variables, using a partial F-test on the inclusion of markers (or change in markers) in the model at $\alpha$ = .05 and test the sensitivity of the result to the inclusion of confounders and mediators. If predictors are strongly correlated ($\rho$ of IL-6 and MCP-1 in preliminary data is .93), we will assess individual predictor level effects in models stratified by predictor, again adjusting for aforementioned covariates and testing effect estimates at $\alpha$ = .05, making adjustments for a higher type 1 error rate due to multiple comparisons. 

**Aim 2:**

For Aim 2, we will include amyloid deposition measured as mean SUVR and its interaction with inflammatory markers to the significantly predictive models in Aim 1, adjusting for aforementioned covariates and testing result sensitivity to confounding, using analogous testing methods and $\alpha$ levels to Aim 1, to test whether the interaction's inclusion provides evidence of decline in outcomes far and above that shown in Aim 1. If none of the models showed an effect in Aim 1, we will consider including interaction to all of the predefined models. Converting amyloid deposition presence into a binary variable will be considered and subsequent models assessed at the discretion of the investigator.

**Power:**

The given sample size (N=175) has adequate power to detect correlations in the range calculated from preliminary data, where we were able to compute the correlations between cortical thickness, CVLT-CNG3, IL-6 and MCP-1. Power is calculated for a correlation test between two continuous variables since our tests will be linear regression tests. We will conduct omnibus testing to avoid unnecessary secondary testing in models lacking predictive significance ($\alpha$ = .05). From the table, modest p-value adjustments ($\alpha$ = .01) can be made to account for type 1 error while still preserving power for the range of correlations, and conservative adjustments, e.g. Bonferroni, ($\alpha$ = .0001) still allow us to detect moderately strong correlations ($\rho$ = .35 implies power of .82 with given sample size when $\alpha$ = .0001). Power estimates are preliminary and may be increased with the addition of precision variables or weakened by the inclusion of confounders.


**MLR Power**
**2 sample correlation test**
**effect of confounders on power**
\scriptsize
```{r table, echo = FALSE, warning=FALSE}

# Load in the preliminary data:

memdat <- read.csv("~/Downloads/PrelimData.csv")

#find correlations for each variable pair
mem_corr <- cor(memdat)

#correlations for each outcome and predictor pair
corr_cvlt_il6  <-  mem_corr[1,3]
corr_cvlt_mcp1 <-  mem_corr[1,4]
corr_cort_il6  <-  mem_corr[2,3]
corr_cort_mcp1 <-  mem_corr[2,4]

corr_range_vec <- c(corr_cvlt_il6, corr_cvlt_mcp1, corr_cort_il6, corr_cort_mcp1)
library(powertools)
tabmat1 <- matrix(c(corr.1samp(N= 175, rhoA = corr_range_vec[1], alpha = .05, sides = 2),
                    corr.1samp(N= 175, rhoA = corr_range_vec[2], alpha = .05, sides = 2),
                    corr.1samp(N= 175, rhoA = corr_range_vec[3], alpha = .05, sides = 2),
                    corr.1samp(N= 175, rhoA = corr_range_vec[4], alpha = .05, sides = 2),
                    corr.1samp(N= 175, rhoA = corr_range_vec[1], alpha = .01, sides = 2),
                    corr.1samp(N= 175, rhoA = corr_range_vec[2], alpha = .01, sides = 2),
                    corr.1samp(N= 175, rhoA = corr_range_vec[3], alpha = .01, sides = 2),
                    corr.1samp(N= 175, rhoA = corr_range_vec[4], alpha = .01, sides = 2),
                    corr.1samp(N= 175, rhoA = corr_range_vec[1], alpha = .0001, sides = 2),
                    corr.1samp(N= 175, rhoA = corr_range_vec[2], alpha = .0001, sides = 2),
                    corr.1samp(N= 175, rhoA = corr_range_vec[3], alpha = .0001, sides = 2),
                    corr.1samp(N= 175, rhoA = corr_range_vec[4], alpha = .0001, sides = 2)), byrow = TRUE, ncol = 4)
rownames(tabmat1) <- c("alpha = .05", "alpha = .01", "alpha = .0001")
colnames(tabmat1) <-c("Rho = -.26", "Rho = -.32", "Rho = -.60", "Rho = -.69")

kableExtra::kable(tabmat1)

correlations <- seq(0.25, 0.70, by = 0.05)
alphas <- c(0.05, 0.01, 0.5/60)
alpha_names <- c("Power, alpha = .05", "Power, alpha = .01", "Power, alpha = .5/60")
 
# Build table
results <- data.frame(
  Correlation = paste0("Correlation = ", correlations),
  check.names = FALSE
)
 
for (i in seq_along(alphas)) {
  col_vals <- sapply(correlations, function(rho) {
    corr.1samp(N = 175, rhoA = rho, alpha = alphas[i], sides = 2)
  })
  results[[alpha_names[i]]] <- col_vals
}
 
print(results, row.names = FALSE)
```


```{r chunk2, echo = FALSE, warning=FALSE}

mod_ftest1 <- lm(CVLT_CNG3 ~ IL_6+MCP_1, data = memdat)
summary(mod_ftest1)

mod_ftest2 <- lm(CORT_CNG3 ~ IL_6+MCP_1, data = memdat)
summary(mod_ftest2)
#.48, .113

mlrF.overall(
N = 175,
p = 4,
Rsq = .48,
fsq = NULL,
alpha = 0.05,
power = NULL,
random = FALSE,
v = FALSE
)

mlrF.overall(
N = 175,
p = 4,
Rsq = .113,
fsq = NULL,
alpha = 0.05,
power = NULL,
random = FALSE,
v = FALSE
)

correlations <- seq(0.10, 0.50, by = 0.05)
 
results <- data.frame(
  Correlation = paste0("Correlation = ", correlations),
  "Power for Overall F, 4 predictors" = sapply(correlations, function(x) {
    mlrF.overall(
      N      = 175,
      p      = 4,
      Rsq    = x,
      fsq    = NULL,
      alpha  = 0.05,
      power  = NULL,
      random = FALSE,
      v      = FALSE
    )
  }),
  check.names = FALSE
)
 
print(results, row.names = FALSE)
```

```{r 2samptest, echo = FALSE, warning=FALSE}

# corr.2samp(
# n1 = 150,
# n.ratio = 1/7,
# rho1 = 0,
# rho2 = X,
# alpha = 0.05/16,
# power = NULL,
# sides = 2,
# v = FALSE
# )
# 
# corr.2samp(
# n1 = NULL,
# n.ratio = 1,
# rho1 = NULL,
# rho2 = NULL,
# alpha = 0.05,
# power = NULL,
# sides = 2,
# v = FALSE
# )
# 
# corr.2samp(
# n1 = NULL,
# n.ratio = 1,
# rho1 = NULL,
# rho2 = NULL,
# alpha = 0.05,
# power = NULL,
# sides = 2,
# v = FALSE
# )


rho2_vals <- seq(0.45, 0.70, by = 0.05)
 
# Each ratio gives clean whole-number splits summing to 175:
# n.ratio = 1/6 -> n1 = 150, n2 = 25  (150 + 25 = 175)
# n.ratio = 2/5 -> n1 = 125, n2 = 50  (125 + 50 = 175)
# n.ratio = 3/4 -> n1 = 100, n2 = 75  (100 + 75 = 175)

n1_vals   <- c(150, 125, 100, 75, 50, 25)
n2_vals   <- c(25,  50,  75, 100, 125, 150)
ratios <- n2_vals/n1_vals
col_names <- paste0(

  " (n1=", n1_vals, ", n2=", n2_vals, ")"
)
 
#  "n.ratio = ", c("1/6", "2/5", "3/4", "4/3", "5/2", "6"),
results <- data.frame(
  "rho2" = paste0("", rho2_vals),
  check.names = FALSE
)
 
for (i in seq_along(ratios)) {
  results[[col_names[i]]] <- sapply(rho2_vals, function(x) {
    round(corr.2samp(
      n1      = n1_vals[i],
      n.ratio = ratios[i],
      rho1    = 0,
      rho2    = x,
      alpha   = 0.05,
      power   = NULL,
      sides   = 2,
      v       = FALSE
    ),3)
  })
}
 
print(results, row.names = FALSE)

kableExtra::kable(results)
```