---
header-includes:
  - \usepackage{xcolor}
  - \usepackage{colortbl}
  - \usepackage{caption}
  - \captionsetup[table]{font=scriptsize}
fontsize: 12pt
linestretch: 2
output:
  pdf_document: default
  html_document: default
---
```{r table, echo = FALSE, warning=FALSE}
source("~/Documents/GitHub/BIOS6624/P2/Code/P2_Power_Analysis.R")
```

**Jacob Kebe**

**Statistical Analysis Plan and Power Analysis for Bettcher Grant**

**Aim 1:** We will consider outcomes as the change in each of the variables in table 3, as well as cortical thickness, over 1 year. Using multiple linear regression (MLR), we will model these continuous outcomes predicted by all inflammatory markers - (IL-6, TNF- alpha,MCP-1,Eotaxin-1) (all continuous) at baseline and as their difference over a year, f or a total of 8 predictors. Unless model assessment shows poor fit, we will assume linear relationships between outcomes and predictors. We will explore variable distributions to see whether data transformation (e.g. log-transforming right-skewed variables) is necessary. Missing values and outliers will be flagged and potentially removed, for example if an observation has multiple markers with values further than 2 SDs from the mean. We will assess covariates (age, sex, CV risk, immunological health history conditions, APOE genotype, baseline outcome values) and determine whether they are confounders (causally affecting the predictors and outcomes), precision variables (causally affecting the outcome), or mediators (variables through which the predictor impacts the outcome), using both domain knowledge and by assessing correlation structure, regression coefficients and significance. We will check whether covariates are highly correlated with each other ($\rho$ > .7). If two precision variables are strongly correlated, only the more predictive one will be kept in the model. Because predictors are highly correlated, and cognitive outcomes are also expected to be highly correlated, we will make the modest $\alpha$-level adjustment of .05/2 to account for the two outcome categories (cortical thickness and memory measures) we will be testing, because preliminary data shows a weak correlation between cortical thickness and the memory measure CVLT ($\rho$ = .16). If the cognitive outcomes show an unexpected correlation structure we will consider alternative corrections, such as correcting for the total number of outcomes ($\alpha$ = .05/6). We will have a separate omnibus test model for 1a. (including all markers) and 1b. (including change in all markers). We will test the overall effect of the predictors  on each outcome in the presence of age, sex, and precision variables, using a partial F-test (DF=4) on the inclusion of markers (or change in markers for 1b) in the respective MLR model at $\alpha$ = .025 and test the sensitivity of the result to the inclusion of confounders and mediators. We will only test individual level effects for each predictor for outcomes whose partial F-test was significant. Because predictors are strongly correlated ($\rho$ of IL-6 and MCP-1 in preliminary data is .93), we will assess individual predictor level effects in models stratified by predictor, again adjusting for aforementioned covariates and testing effect estimates using a t-test on the predictor's beta coefficient at $\alpha$ = .025. 

**Aim 2:** For Aim 2, we will add amyloid deposition (AD) measured as mean SUVR and its interaction with inflammatory markers (or their change over a year) to the significantly predictive MLR models in Aim 1, adjusting for aforementioned covariates and testing result sensitivity to confounding, using analogous testing methods and $\alpha$ levels to Aim 1, to test whether the interaction's inclusion provides evidence of decline in outcomes far and above that shown in Aim 1. If none of the models showed an effect in Aim 1, we will consider including interaction to models for all outcomes. For each outcome, we will first conduct an omnibus test of AD's interaction with all predictors, again using a partial F test on the MLR with 4 DF. We will then again test individual models for each predictor's interaction with AD adjusting for the predictor, AD, and covariates, using a t-test on the interaction beta coefficient.  We will then check the sensitivity of the results to the inclusion of confounders. Converting amyloid deposition presence into a binary variable will be considered and subsequent models assessed at the discretion of the investigator.

**Power:**  For power calculations, we assume a power of .8 is adequate to conduct a valid study. We will conduct omnibus testing to avoid unnecessary secondary testing in models lacking predictive significance ($\alpha$ = .025).   We estimate the power for the partial F tests for omnibus testing using an overall F-test for MLR power calculation, using a range of $R^2$ values from linear models of CVLT and cortical thickness as functions of IL-6 and MCP-1 from the preliminary data ($R^2$=.11,.48), using DF=4. We have adequate power for the range of R squared values at $\alpha=.05/2,.05/6$.
```{r rsqtable, echo = FALSE, warning=FALSE}
kable(results_rsq,  format = "latex", escape = FALSE, booktabs = TRUE,
      caption = "\\scriptsize Power to detect F-test significance using alpha = .05/2 and .05/6 and range of R squared values from preliminary data models.") %>%
  kable_styling(latex_options = c("hold_position", "scale_down"),
                font_size = 8)
```
 For individual effect tests, power is calculated for a correlation test between two continuous variables since our tests will be linear regression tests.  The given sample size (N=175) has adequate power to detect correlations in the range calculated from preliminary data at $\alpha=.05/2,.05/6$, where we were able to compute the correlations between cortical thickness, CVLT, IL-6 and MCP-1. Power estimates are preliminary and may be strengthened with the addition of precision variables or weakened by the inclusion of confounders. 
```{r 1samptable, echo = FALSE, warning=FALSE}
kable(results_1samp,  format = "latex", escape = FALSE, booktabs = TRUE, caption = "\\scriptsize Power to detect correlation range from preliminary data correlation matrix using alpha = .05/2, .05/6") %>%
  kable_styling(latex_options = c("hold_position", "scale_down"),
                font_size = 8)
```

To power Aim 2, we will make the conservative assumption that AD presence is binary, and conduct a 2 sample correlation power calculation, using the correlation range from preliminary data as a benchmark for correlation differences between the AD presence group and the absence group. This will be an underestimation of the power to detect a continuous on continuous interaction with 1 sample.  From the tables, we have adequate power when N=175 to detect moderate to strong correlation differences, especially when the 2 samples are closer in size, under $\alpha=$ .05/2,.05/6. Code for power analyses is available at https://github.com/jkebe2002/BIOS6624/tree/d71f7df370b60dc3480000b623bcf63ebab3e679/P2. Preliminary data used for calculations is stored locally at C:/Users/jkebe/Downloads/PrelimData.csv
```{r 2samptable, echo = FALSE, warning=FALSE}
kable(results_display, format = "latex", escape = FALSE, booktabs = TRUE, caption = "\\scriptsize Power to detect correlation differences between two samples of varying sizes where N=175 and alpha = .025. Values above .8 are bolded.") %>%
  kable_styling(latex_options = c("hold_position", "scale_down"),
                font_size = 9)
```
\newpage

```{r 2samptable2, echo = FALSE, warning=FALSE}
kable(results_display2, format = "latex", escape = FALSE, booktabs = TRUE, caption = "\\scriptsize Power to detect correlation differences between two samples of varying sizes where N=175 and alpha = .05/6. Values above .8 are bolded.") %>%
  kable_styling(latex_options = c("hold_position", "scale_down"),
                font_size = 9)
```

\scriptsize
