---
title: "HAART Treatment Effectiveness in Drug Users: 2-Year Study"
author: "Jacob Kebe"
output: pdf_document
documentclass: article
geometry: margin=1in
fontsize: 12pt
linestretch: 2
header-includes:
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage{multirow}
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{dcolumn}
  - \usepackage{fvextra}
  - \usepackage{bm}
---



```{r source, echo=FALSE,warning=FALSE,message=FALSE}
library(tidyverse)
library(dplyr)
library(cmdstanr)
library(bayesplot)  # diagnostic plots of the MCMC chains
library(posterior)  # for summarizing posterior draws
library(bayestestR) # for calculating higest density posterior intervals
library(mcmcse)     # for calculating MCMCSE's
library(loo)        # for getting model fit statistics (WAIC and LOO-IC)
library(dplyr)
library(tibble)
library(ggplot2)
library(gridExtra)
#NOTE: USE THIS TO PRODUCE hivdat_clean data!
source("~/Documents/GitHub/BIOS6624/P1/Code/P1_Data_Management.R")
#Use this to produce final report tables graphs etc!
source("~/Documents/GitHub/BIOS6624/P1/Code/P1_Final_Analysis_Code.R")
```

# Introduction
Human Immunodeficiency Virus (HIV) is a devastating disease that attacks the body's immune system. Highly Active Antiretroviral Treatment is the standard treatment for those infected. Using data collected from the Multicenter AIDs Cohort Study, with up to 8 years of data collected on a prospective cohort of homosexual and bisexual men infected with AIDs and treated regularly with HAART, we are interested in whether outcomes after two years differ for hard drug users and those that do not use hard drugs. We are interested in whether the change from baseline to 2 years of our 4 outcomes of interest (HIV copies / ml blood, CD4+ T cell count, SF-36 Quality of Life Scores, Mental and Physical) differs significantly when comparing hard drug users to non-hard drug users, as well as whether this difference can be explained by self-reported adherence at 2 years. Variables of interest recorded in the data include but are not limited to hard drug use, the aforementioned outcomes, age, race, education level, BMI, and smoking status. Other covariates, such as lab values, are partially available with high rates of missingness.
\footnotesize
```{r tableone, echo=FALSE,warning=FALSE,message=FALSE}
library(gtsummary)
t1
```
\normalsize

# Methods
Viral load (HIV copies / ml blood) was log transformed due to being nonnegative and highly right skewed. We filtered data to data for years 0 and 2. We considered patients who had all outcomes measured at the two time points in order to draw inference on the same population for each outcome, reducing the sample size from 715 to 476.
Data is assumed to be missing completely at random, and complete case analysis is utilized. We redefined variable missingness using the missing definitions in the data dictionary. We looked at missingness rates of variables against factors of the outcomes and primary predictor, and found no concerning trends (Appendix). For example, we saw similar rates of missingness across values of CD4+ T cell count, so we were not concerned about data being missing due to poor immunological conditions. There was also not a strong missingness pattern over time, so dropout was not a concern especially in the first two years. For complete case analysis we dropped 13 participants with missing baseline BMI, resulting in a final sample size of 463. BMI values higher than 250 or less than 0 were considered missing. 

```{r ggplots, echo=FALSE,warning=FALSE,message=FALSE}

####JACOB READ THIS LATER
####SUPPRESS CHAIN MESSAGES
####FONT SIZE OF TABLES
####GGPLOT GRAPHS
#### TABLE 1 REDUCE FONT
#### CAPTIONS FOR KABLE TABLEs
####CAPTIONS FOR GGPLOT

library(gridExtra)
library(grid)
library(egg)

bottom_text <- textGrob(
  "Figure 1: Outcome Densities stratified by drug use at baseline and 2 years.",
  gp = gpar(fontfamily = "serif", fontsize = 8),
  hjust = 1, # Horizontal justification
  x = 1        # X position (adjust as needed)
)
grid.arrange(grobs = lapply(list(p_lvl0, p_leu3n0, p_aggment0, p_aggphys0, p_lvl2, p_leu3n2, p_aggment2, p_aggphys2),set_panel_size,
    width = unit(.9, "in"),
    height = unit(1.5, "in")), widths =c(1,1,1.1,1.2), ncol=4, bottom = bottom_text) 
```
We calculated difference in log viral load, CD4+ T cell count, and the quality of life scores from baseline to year 2 as the outcomes of interest. We considered baseline measures for hard drug use and other predictors in order to have pre-treatment values to adjust for, and considered adherence at year 2. Race, education, and adherence were aggregated into 2 categories each due to imbalanced proportions across categories and to avoid inflating the number of model parameters. We used non-Hispanic White and other for race, college degree and higher vs. none for education, and 95% adherence or better vs. worse as the respective binary categories, at the suggestion of the investigator.
The primary research question was addressed using multiple linear regression in both a Bayesian and frequentist setting, with outcomes as a function of hard drug use, adjusting for baseline age, sex, BMI, education level, and race, as these were the standard variables to adjust for according to the investigator. Bayesian analysis assumes that the effect estimate has a distribution. For prior distributions of the effects of our predictors, we used noninformative normal priors with mean 1 for the intercept and 0 for the predictors, with standard deviation 100. The prior for the model's standard deviation was a half-normal with mean 0 and standard deviation 100. We assessed the mean of the posterior distribution of the estimate for the effect of hard drug use for the four outcome models, as well as the 95% highest posterior density interval and posterior probability of a clinically meaningful difference to answer the questions of interest. Clinically meaningful differences, according to the investigator, were .5 log copies HIV per ml blood for log viral load, 50 cells/mm$^3$ for CD4+ T cell count, and 2 points for the quality of life measures. Bayesian estimates and error measures were compared to frequentist estimates to assess the validity of the Bayesian approach for the 4 main outcome models. We also fit Bayesian multiple linear regression models excluding hard drug use for model fit assessment, using change in the model's Leave-One-Out Cross-Validation Information Criterion (LOO-IC), a Bayesian measure of model predictive accuracy. To study the effect of adherence, we modeled the outcomes against drug use in addition to adherence, adjusting for aforementioned covariates. Therefore we fit 16 models total. We used cmdstan and base R linear modelling functions for Bayesian and frequentist approaches respectively. Data is stored locally at C:/Users/jkebe/Downloads/hiv_6624_final.csv on Jacob Kebe's laptop.

# Results
\scriptsize
```{r bayestable, echo=FALSE,warning=FALSE,message=FALSE}
knitr::kable(bayes_table_matrix, caption="Posterior Means and Intervals for Drug Use Effect. Posterior Probabilities are probability of clinically meaningful treatment effect.")
```
\normalsize
Results come from models adjusted for age, sex, race, education, and BMI. For non users, we found that the viral load after 2 years was 9.08% (95% HPDI:[3.82%,20.97%]) that of baseline. We found that for hard drug users, the viral load after 2 years was 8.82% [3.55%,21.63%]. On the log scale, drug users' two year difference in log viral load was .03 (95% CI: -.40,.47]) less than that of non-drug users. The posterior probability that drug users' difference in log viral load differed from that of non-drug users was .025, providing evidence of no meaningful difference. Including hard drug use in the model did not improve the LOO-IC.
The average treatment effect in CD4+ T cell counts for non-users was an increase of 103.27 (95% HPDI: [.119,210.761]) after 2 years, a significant change. The difference in CD4+ T cell counts from year 0 to year 2 for drug users was on average 147.47 (95% CI: [88.13,207.00]) less than non-drug users, effectively eliminating the effect of treatment. Average difference in T cell count for users was -44.196 (95% HPDI: [-161.085,70.717]). The posterior probability that this discrepancy is over 50 T cells/mm$^3$ is .999, providing evidence that treatment was less effective for the drug user group than the non drug users for increasing T-cell count. The LOO-IC for the model including hard drugs was an improvement of 23 points over the null model.
The average mental quality of life score for non-users at 2 years was on average 6.839 points higher than baseline (95% HPDI: [-1.546, 15.325]) which was insignificant. Hard drug users' quality of life increase by an average of an additional 1.18 points (95% HPDI: [-3.246,5.305]) which had a probability of .419 of differing by 2 points or more from non-users. No significant effect for either group was found. The LOO-IC did not improve when including hard drug use in the mental QOL outcome model.
The average physical  quality of life score for non-users at 2 years was on average .604 points lower than baseline (95% CI: [-5.024,6.325]) which was not significant. The difference for users was 3.160 points lower (95% CI: [.268,6.036]), with posterior probability of .785 that this differed by more than two points from the non-user trend. The average change in physical quality of life score for users was -3.718 (95% HPDI: [-10.068, 2.692]) which was not significant. The LOO-IC did not improve when including hard drug use in the physical QOL outcome model.
\scriptsize
```{r bayesfreqtable, echo=FALSE,warning=FALSE,message=FALSE}
kableExtra::kable(bayes_vs_freq_matrix,caption="Frequentist Estimates and CI's vs. Bayesian Estimates and HPDI's for Drug User Effect Difference")
```
\normalsize
We compared Bayesian model estimates to models using non-Bayesian multiple linear regression and found no meaningful differences. The frequentist estimates and confidence intervals were directionally equivalent and showed comparable effect size and variability.
We compared the effect estimates for hard drug use on all of the outcomes adjusting for high adherence vs. low adherence (95% and above as high). None of the effect estimates or HPDI's changed sign. For the log viral load model there was .464 probability that the estimates were meaningfully different, for the CD4+ model there was .489 probability, for Mental QOL score there was .529 probability, and for Physical QOL there was .492 probability. For no outcome was it shown that there was substantial evidence of adherence significantly mediating the effect of hard drug use. In summary, log viral load decreased significantly for both groups, and CD4+ T cell count increased in the non-user group, an effect not seen in the user group. Model diagnostics for all models including trace plots, autocorrelation plots (both in appendix), and Gelman-Rubin statistic showed no issues with model fit.
\scriptsize
```{r adhtable, echo=FALSE,warning=FALSE,message=FALSE}
kableExtra::kable(noquote(adh_tab), caption="Estimates and HPDI's comparing models with and without adherence variable.")
```
\normalsize
# Conclusions / Discussion

The prevalance of hard drug use makes it imperative to study how it may affect treatment utility, especially HAART, an integral intervention against HIV, a destructive and pervasive disease. We found that in a 2 year period, HAART effectively reduced HIV copies per ml blood for users and non-users of hard drugs. We also found that treatment was effective in promoting CD4+ T cell count for non-users, but that this effect was not seen in drug users. Adherence did not play a significant role in the effects of hard drugs on the outcomes. This analysis provides clinicians with evidence of HAART's effectiveness in reducing HIV presence, and calls upon further research into how treatment for hard drug users can better improve immunological response. This study was limited by missingness and the assumptions required for complete case analysis, as well as the difference in non-user (n=437) vs. user groups (n=39).








\newpage

# Appendix

# Trace Plots for Outcome Models

We include trace plots for the beta term for treatment effect on non-users, additional effect for users, and sigma, for the four outcome models (Log viral load, T cell count, mental and physical QOL score).

```{r traceplots, echo=FALSE, warning=FALSE,message=FALSE}
draws1 <- as_draws_array(fit_lvload$draws())  # dimensions: iterations x chains x parameters
draws_df1 <- as_draws_df(fit_lvload$draws())  # tidy format for bayesplot / ggplot

draws2 <- as_draws_array(fit_leu3n$draws())  # dimensions: iterations x chains x parameters
draws_df2 <- as_draws_df(fit_leu3n$draws())  # tidy format for bayesplot / ggplot

draws3 <- as_draws_array(fit_agg_ment$draws())  # dimensions: iterations x chains x parameters
draws_df3 <- as_draws_df(fit_agg_ment$draws())  # tidy format for bayesplot / ggplot

draws4 <- as_draws_array(fit_agg_phys$draws())  # dimensions: iterations x chains x parameters
draws_df4 <- as_draws_df(fit_agg_phys$draws())  # tidy format for bayesplot / ggplot
params <- c("beta[1]", "beta[2]", "sigma")

# Trace plots
# Trace plots show whether chains are mixing well and exploring the parameter space.
# Chains should mix well (lines overlap)
# No long trends (no “stickiness”)
# Ideally all chains overlap around the same mean

mcmc_trace(draws1, pars = params)
mcmc_trace(draws2, pars = params)
mcmc_trace(draws3, pars = params)
mcmc_trace(draws4, pars = params)
```

# Autocorrelation Plots for Outcome Models

We include autocorrelation plots for the beta term for treatment effect on non-users, additional effect for users, and sigma, for the four outcome models (Log viral load, T cell count, mental and physical QOL score).

```{r autocorrplot, echo=FALSE, warning=FALSE,message=FALSE}

# Auto-correlation 
# Autocorrelation should decay quickly
# High autocorrelation → may need more iterations or thinning
mcmc_acf(draws1, pars = params)
mcmc_acf(draws2, pars = params)
mcmc_acf(draws3, pars = params)
mcmc_acf(draws4, pars = params)

```


# Table 1 (Non-filtered Dataset)


```{r larget1,echo=FALSE,message=FALSE,warning=FALSE}
#Table 1:

#t1 is your table 1 object.
library(gtsummary)
library(dplyr)

# adherence at year 2
hivdat$ADH <- if_else(hivdat$ADH>2,1,0)
hivdat$ADH_cat <- if_else(hivdat$ADH>2,"95% Adherence or Higher","Less than 95% Adherence")


#education:  5-7 or 1-4 for college educ or higher vs. no college
hivdat$educ_cat <- if_else(hivdat$EDUCBAS>4, "College Degree or Higher", "No College Degree")

# RACE Race 1= White, non-Hispanic
# 2= White, Hispanic
# 3= Black, non-Hispanic
# 4= Black, Hispanic
# 5= American Indian or Alaskan
# Native
# 6= Asian or Pacific Islander
# 7= Other
# 8= Other Hispanic (created
#                    for 2001-03 new recruits)
# Blank= Missin

#create race cat. variable with two options

hivdat$race_cat <- if_else(hivdat$RACE==1, "White, Non-Hispanic", "Other")

# SMOKE Smoking status 1= Never smoked
# 2= Former smoker
# 3= Current smoker
# Blank= Missing
hivdat$smoke_cat <- if_else(hivdat$SMOKE==1, "Never Smoked", 
                                  if_else(hivdat$SMOKE==2, "Former Smoker", 
                                          if_else(!is.na(hivdat$SMOKE), "Current Smoker", NA)))



# Convert to factor if needed
hivdat <- hivdat %>%
  mutate(
    hard_drugs = factor(hard_drugs)
  )

# Create gtsummary Table 1
t1_large <- hivdat %>%
  select(
    hard_drugs,
    VLOAD,
    LEU3N,
    AGG_MENT,
    AGG_PHYS,
    educ_cat,
    race_cat,
    smoke_cat,
    BMI,
    age,
    ADH_cat
  ) %>%
  tbl_summary(
    by = hard_drugs,
    label = list(
      VLOAD   ~ "log(HIV copies per mL of blood)",
      LEU3N   ~ "CD4+ T cell Count",
      AGG_MENT ~ "Mental Quality of Life Score",
      AGG_PHYS ~ "Physical Quality of Life Score",
      educ_cat  ~ "Education Level",
      race_cat  ~ "Race",
      smoke_cat ~ "Smoking Status",
      BMI       ~ "Body Mass Index (BMI)",
      age       ~ "Age",
      ADH_cat   ~ "Adherence"
    ),
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = all_continuous() ~ 2,
    missing = "ifany"   # <-- key change
  ) %>%
  modify_header(label = "**Variable**") %>%
  modify_caption("**Table of Descriptive Statistics by Drug Use (Including Missingness)**") %>%
  bold_labels()

t1_large

```


# Missingness by Outcome and Treatment 


```{r missingness, echo=FALSE}
library(naniar)
library(tidyverse)

hivdat$ment_cat <- cut(hivdat$AGG_MENT, breaks = 10)
hivdat$phys_cat <- cut(hivdat$AGG_PHYS, breaks = 10)
hivdat$LEU3N_cat <- cut(hivdat$LEU3N, breaks = 10)
hivdat$VLOAD_cat <- cut(hivdat$VLOAD, breaks = 10)
gg_miss_fct(hivdat[hivdat$years ==0,], ment_cat)
gg_miss_fct(hivdat[hivdat$years ==0,], phys_cat)
gg_miss_fct(hivdat[hivdat$years ==0,], LEU3N_cat)
gg_miss_fct(hivdat[hivdat$years ==0,], VLOAD_cat)
gg_miss_fct(hivdat[hivdat$years <3,], hard_drugs)
gg_miss_fct(hivdat, years)

```



 